Hypertension and microbiome

### 2019-04-25 articletwo.Rmd
Minor

    if("clean" %in% names(args)) {
        unlink("rds", recursive=TRUE)
        unlink("cache", recursive=TRUE)
    }
    if ("tags" %in% names(args)) 
        rtags('~/phd/research/articletwo/', recursive = FALSE, pattern = '\\.[RrSs](rw|md)?$',ofile = '~/phd/research/articletwo/TAGS', verbose = TRUE, append = FALSE)
}
```

@@

```{r test scaling}
meta(pseq.genus) %>% summarize(mean(MAP), sd(MAP), mean(SYSTM), sd(SYSTM), mean(DIASM), sd(DIASM), mean(PULSEPRESSURE), sd(PULSEPRESSURE)) %>% kable
```

@@

```{r maaslin runs, results = 'hide'} 
maaslin <- list()

@@

```{r Write figure 2}
fig2head <- "Tenth of bacterial variation can be explained by two first principal axes"
fig2foot <- sprintf("Principal component analysis for bacterial abundances at species level. Centered log ratio transform is performed for the compositional data. First two axes explain %.1f%% of the variation. Research participants with hypertension are denoted by red dot.", sum(pca.percentage[1:2]))
writeimage(doc, 2, "cache/pca.png", fig2head, fig2foot)
```

@@

```{r maaslin}
maas.results.df.signf <- maaslin$genus %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
           beta = Coefficient) %>%
   dplyr::filter(Qval < 0.05)

maaslinplot <- plot_maaslin_heatmap_final(maas.results.df.signf %>% dplyr::filter(Variable != "HYPERTENSION"),
                      beta.maxabs=max(abs(maas.results.df.signf[,'beta'])),
                      order.by.category=FALSE)

@@

## Figure 4

```{r PC plot}
pca.prcomp.df <- pca.prcomp$x[,1:5] %>% data.frame
#pca.prcomp$sdev[1:20]^2/sum(pca.prcomp$sdev^2)
#cumsum(pca.prcomp$sdev[1:20]^2/sum(pca.prcomp$sdev^2))

X <- base::merge(meta(pseq.species), pca.prcomp.df, by=0, all=TRUE) %>%
    dplyr::rename(Sample_ID = Row.names)

X.res <- lapply(vectortolist(paste0("PC", seq(5))), function(axis) {
    modelstring <- paste0("%s ~  BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + DIAB + BL_USE_RX_C03 + BL_USE_RX_C07 + BL_USE_RX_C08 + BL_USE_RX_C09 + ", axis)
    calculateglm(X, modelstring = modelstring, filterstr = "PC")
})

Z.df <-  do.call(rbind.data.frame, X.res) %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="response", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(p.value, method="BH"),
                  beta = estimate,
                  Feature = term)

p2 <- plot_maaslin_heatmap_final(Z.df , fname="Feature", pname="Name",
                                  xlab.text = "PC", signif="Qval", cluster=FALSE,
                                  v.title.name = "Effect size (PC)",
                                  ystr="reorder(Name, Qval)", sign.name="Significance.PC",
                                  no.sign.legend=TRUE)

ggsave(file = "cache/pca.p2.png", plot = p2, height = 8, width = 8, dpi = 300) 
```

```{r Write figure 4}
fig4head <- "P2"
fig4foot <- ""
writeimage(doc, 4, "cache/pca.p2.png", fig4head, fig4foot)
```

## Figure 5

```{r PCA heatmap}
pca.axis <- pca.prcomp$x %>% data.frame
pca.dset <- base::merge(meta(pseq.species), pca.axis[,1:5], by=0, all=TRUE) %>%
  dplyr::rename(Sample_ID = Row.names)

pca.dset %>% head

pca.ret <- lapply(paste0("PC", seq(1,5)), function (pc) {
    calculateglm(pca.dset,
                 modelstring = paste("%s ~ BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + DIAB + BL_USE_RX_C03 + BL_USE_RX_C07 + BL_USE_RX_C08 + BL_USE_RX_C09  +", pc),
                 filterstr = "PC") %>%
        select(response, estimate, p.value) %>%
        mutate(model_name = pc)
})

pca.ret  %>%
    data.table::rbindlist() %>%
    as.data.frame %>%
    mutate(p.value = p.adjust(p.value, method="BH")) %>%
    arrange(p.value) %>%
    kable
```

## Loadings for principal component axis 3 (Figure 5)

```{r Main contributors of PC3}
pca.rotation.abs <- abs(pca.prcomp$rotation)
pca.loading <- sweep(pca.rotation.abs, 2, colSums(pca.rotation.abs), "/") %>%
	    as.data.frame %>%
    tibble::rownames_to_column(var = "term") %>%
    mutate(name = rename.genus(term))

pca.loading.pc3.top20 <- pca.loading %>%
    select(name, PC3) %>%
    arrange(-PC3) %>%
    head(20)

pca.pc3.gg <- ggplot(data=pca.loading.pc3.top20, aes(x=reorder(name, -PC3), y=100*PC3)) +
    geom_bar(stat="identity", width=0.5, position = position_dodge(preserve = "single")) +
    theme_classic(20) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.3, hjust = 1)) +
    xlab("") +
    ylab("Contribution (%)") +
    scale_y_continuous(limits=c(0, 2.7), oob = rescale_none, expand = c(0,0)) +
    theme(axis.title.x=element_blank(),
          text = element_text(size=20),
          legend.justification=c(1,0),
          legend.position=c(1.03, 0.75),
          legend.title=element_text(size = 20),
          legend.direction="vertical",
          legend.text=element_text(size = floor(0.75*20)),
          legend.box.margin = margin(1, 1, 1, 1, "mm"),
          legend.box.background = element_rect(colour = "black"),
          aspect.ratio = 0.7)
   ggsave(file = "cache/pca.pc3.png", plot = pca.pc3.gg, height = 8, width = 8, dpi = 300) 
```

```{r Write figure 5}
fig5head <- "Bacteria causing dysentery and diarrhea contribute to third principal component"
fig5foot <- sprintf("Top 20 bacterial contributors to third principal coordinate at species level. Centered log ratio transform is performed for the compositional data. Third principal component explain %.1f%% of the total variation. Plasmid DNA is denoted using star.", pca.percentage[3])
writeimage(doc, 5, "cache/pca.pc3.png", fig5head, fig5foot)
```

@@

```{r lactobacillus table}
allruns.unamb <- turkumetabolites::unambiguous.lmrank(dset$fr02, linear.lmrank)
allruns.sup <- turkumetabolites::pub.lmrank(allruns.unamb,  p = 0.05/330)

maaslin.lacto.max.table <- maaslin$lacto.max %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  beta_ci =sprintf("%.1E", Coefficient, NA),
                  p.value = sprintf("%.2f", p.adjust(P.value, method="BH"))) %>%
    merge(., names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    select(Feature, Name, beta_ci, p.value) %>%
    gather(variable, value, -Feature, -Name) %>%
    unite(temp, Name, variable) %>%
    spread(temp, value) %>%
    mutate(Feature = rename.genus(gsub("_Bacteria", "", Feature))) %>%
    arrange(Feature)

>>>>

    if("clean" %in% names(args)) {
        unlink("rds", recursive=TRUE)
        unlink("cache", recursive=TRUE)
    }
    if ("tags" %in% names(args)) 
        rtags('~/phd/research/articletwo/', recursive = FALSE, pattern = '\\.[RrSs](rw|md)?$',
              ofile = '~/phd/research/articletwo/TAGS', verbose = TRUE, append = FALSE)
}
```

@@

@@

```{r maalin manual, eval = FALSE}
test.abud <- microbiome::abundances(microbiome::transform(pseq.genus.core, "compositional")) %>%
    t %>%
    as.data.frame
df <- test.abud %>% mutate_all(funs(asin(sqrt(.))))
```

```{r maaslin runs, results = 'hide'} 
maaslin <- list()
lm (Sutterella_Bacteria ~ MAP + BL_AGE + SEX + BMI + CURR_SMOKE + Q57X + DIAB + BL_USE_RX_C03 + 
                           BL_USE_RX_C07 + BL_USE_RX_C08 + BL_USE_RX_C09, )

@@

```{r Write figure 2}
fig2head <- "Principal component analysis for bacterial abundances"
fig2foot <- sprintf("Principal component analysis for bacterial abundances at species level. Centered log ratio transform is performed for the compositional data. First two axes explain %.1f%% of the variation. Centered log-ratio (CLR) tranformation is used for microbiome abundances.", sum(pca.percentage[1:2]))
writeimage(doc, 2, "cache/pca.png", fig2head, fig2foot)
```

@@

```{r maaslin}
maas.results.df.signf <- maaslin$genus %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  Feature = gsub("_Bacteria", "", Feature),
                  beta = Coefficient) %>%
    dplyr::filter(Qval < 0.05) %>%
    arrange(Feature)

maaslinplot <- plot_maaslin_heatmap_final(maas.results.df.signf %>% dplyr::filter(Variable!="HYPERTENSION"),
                      beta.maxabs=max(abs(maas.results.df.signf[,'beta'])),
                      order.by.category=FALSE)

@@

@@

```{r lactobacillus table}
maaslin.lacto.max.table <- maaslin$lacto.max %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  beta_ci =sprintf("%.1E", Coefficient, NA),
                  p.value = sprintf("%.2f", p.adjust(P.value, method="BH"))) %>%
    merge(., names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    select(Feature, Name, beta_ci, p.value) %>%
    gather(variable, value, -Feature, -Name) %>%
    unite(temp, Name, variable) %>%
    spread(temp, value) %>%
    mutate(Feature = rename.genus(gsub("_Bacteria", "", Feature))) %>%
    arrange(Feature)
